\large How resource abundance and stochasticity affect organisms' space-use requirements

### 2023-09-12 writing/manuscript.Rmd
addressed Mike's comments

---
title: \large How resource abundance and stochasticity affect organisms' space-use requirements
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Simon Wood
    institute: uofe
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: E. Patrícia Medici
    institute: [tapir, iucn, escas]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - uofe: School of Mathematics, James Clerk Maxwell Building, University of Edinburgh, Edinburgh, Unted Kingdom.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'hr-resource-stoch.bib'
csl: 'the-american-naturalist.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{wrapfig} # for wrapping text around figures
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # @reed_frequency_2003 has unicode character U+0308
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
editor_options: 
  chunk_output_type: console
---

@@

@@

knitr::knit_hooks$set(plot = function(x, options) {  # set the new plot hook ...

@@

  # if wrapfigure options are given, create the new opening string for the wrapfigure
  # environment using custom figure alignment and text width
  if(!is.null(options$wrapfigure)) {

@@

@@

\noindent The amount of space organisms use is thought to be tightly linked to the availability of resources within their habitats, such that organisms living in productive habitats generally require less space than those in resource-poor habitats. This hypothesis has widespread empirical support, but existing studies have focused primarily on responses to the *mean* amount of resources, while responses to the variance around the mean are still largely unknown. This is not a trivial oversight. Organisms adjust to variable environmental conditions, so failing to account for the effects of resource unpredictability can result in a limited understanding of organisms' space-use requirements, challenging ecological theory and applied conservation alike. In this study, we first review the literature of studies that estimate animals' space-use requirements as a function of resource abundance. We then leverage this information to provide a unifying framework and hypotheses for the effect of mean *and* variance in resources on organisms' space use. Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resource abundance interact to shape predictable patterns in space use. Finally, we use real-world tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals. Results from the simulations and empirical examples are presented using a fully transparent approach that allows researchers to apply the framework to their own data and inform area-based conservation efforts.

@@

The relationship between organisms' movement and resource abundance has been of interest to biologists for over half a century. @burt_territoriality_1943 considered the search for food as the primary driver for movement within an organism's home range. Three decades after, @southwood_habitat_1977 suggested that change in resource abundance is a strong determinant of how organisms decide where to live and when to reproduce. Two years later, @harestad_home_1979 proposed that the simplest relationship between resource abundance and an organism's home-range size is

@@

\noindent where $H$ is the organism's **home-range size**, $C$ is the organism's resource consumption, and $R$ is the **resources** the organism can access. Harestad and Bunnel's model is simple to conceptualize and allows for testable predictions, but a species' spatial requirements likely depend on numerous other factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020]. While many researchers have since demonstrated that organisms adapt their home ranges in response to resources abundance, few studies build upon others' previous work. Instead, results are reported as independent, novel findings. Consequently, we currently lack a unifying framework for quantifying the effects of resource abundance on organisms' space-use requirements. In addition, while much work has been done on estimating organisms' responses to average resource abundance, there is little information on how they respond to unpredictable changes in resources.

In this paper, we refer to a location's average amount of resources as **resource abundance**, while we use the phrase **resource stochasticity** to indicate the variability in resource abundance due to unpredictable causes. We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an organism might use. For instance, a grassland with relatively low but constant forage availability will require drastically different behaviors and adaptations from a desert location with equally scarce forage but rare, sudden, and strong pulses of resources. In the grassland, an animal may require a large but constant home range as it moves between locations in search of food [*sensu* @teitelbaum_beyond_2019], while an animal in the desert may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are depleted. Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in substantially different movement and life history strategies being selected for. Although it is generally expected that resource unpredictability will decrease organisms' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is little empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].

Here, we illustrate how organisms' space-use requirements depend on both the abundance and unpredictability of resources. First, we set the theoretical background necessary for the successive sections by introducing key concepts and notation. Next, we provide a review of the effects of resource *abundance* on organisms' space-use requirements while suggesting a unifying hypothesis. Afterwards, we provide a review of the effects of resource *stochasticity* on organisms' space-use requirements while suggesting a second unifying hypothesis. Subsequently, we demonstrate the power of these two hypotheses using quantitative, simulated responses to changes in resource abundance and unpredictability. Finally, we demonstrate how this framework can be used in practice to describe the movement ecology of a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado [@medici_movement_2022]. Results from the simulations and empirical example are presented using a fully transparent approach that allows researchers to replicate the work and apply the methods to their own tracking data.

# Simulating resources as a random variable

\noindent In this section, we simulate $R$ using a Gamma distribution with time-dependent (but spatially homogeneous) mean $\mu(t)$ and variance $\sigma^2(t)$, which we write as $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$. Although Gamma distributions are more often parameterized using parameters shape and scale ($k > 0$ and $\theta > 0$) or shape and rate ($\alpha > 0$ and $\beta = 1/\theta > 0$), we use $\e(R) = k \theta$ and $\var(R) = k \theta^2$ to facilitate visualizing the simulations. However, note that $\e(R)$ and $\var(R)$ are not independent because the variance depends strongly on the mean (and vice-versa). As the mean approaches zero (from the positive side, which we write as $\e(R) \rightarrow0^+$) the variance also does: $\e(R) \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \var(R) = 0^+$. This assumption also holds biologically, since resources tend to be less variable when they are less abundant. We assume readers are familiar with the concepts of: random variables, probability distributions, the expected value of a random variable, and the variance of a random variable. We define and explain each of the four concepts in Appendix 1 for readers who need additional information.

@@

```{r mu-hyp, fig.cap="Hypothesized space-use requirements of an organism as a function of resource abundance ($\\e(R)$). We expect low values of $\\e(R)$ to result in high space-use requirements as organisms are forced to explore large areas to collect the resources they require to survive, whether they be range-resident or nomadic. As $\\e(R)$ increases, space-use requirements should decrease nonlinearly until they reach the minimum amount of space required by the organism to survive. Note that the relationship between $\\e(R)$ and space-use requirements cannot be linear because it would require space-use requirements to be negative for high values of $\\e(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/mean-abundance-hr-hypotheses.png')
```

\noindent While organisms' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction. All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001]. Failing to acquire sufficient resources can result in lower fitness, behavioral changes, physiological changes or damage, inability to reproduce, and death. Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism and movement costs [@taylor_energetics_1982]. The hypothesis that space-use requirements decrease with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014]. This is problematic because a linear relationship can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022]. Conceptually, the relationship between space-use requirements and $\e(R)$ must be nonlinear, since there are limits to how much space an organism is able to explore in its lifetime as well as the minimum amount of space it requires to survive [which is necessarily greater than zero -- see figure \@ref(fig:mu-hyp) as well as: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020; and contrast them to estimates based on linear models, e.g.: @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022].

@@

```{r multipanel-constant-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with constant variance (after accounting for changes in the mean). (a.) Resources ($R$, brown-green fill) vary over time and space with a constant the variance (dot color). (b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and constant variance $\\var(R) = \\sigma^2$. The points in the central panel indicate the ralizations of $R$ in panel a.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/habitat-examples-constant-variance.png')
```

It is also worth noting that high $\e(R)$ does not does not necessarily imply that organisms have access to high amounts of resources, as predator avoidance and competition within an organism's habitat will decrease the proportion of $R$ it is able to access. Thus, organisms in habitats with strong competitive pressure and predation may require larger home ranges [@jetz_scaling_2004; @prox_framework_2020], but larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004; but also see @dickie_resource_2022]. Additionally, persistent territorial defense from competitors may prevent organisms from using space freely and as necessary [wolves, *Canis lupus*: @rich_anthropogenic_2012; feral cats, *Felis catus*: @bengsen_feral_2016; Capuchin monkeys, *Cebus capucinus*: @torrez-herrera_monkeys_2020], so the effect of $\e(R)$ on space use may vary between individuals, species, and locations. Still, these pressures simply alter the intensity with which organisms respond to $\e(R)$ but not the shape of the relationship, since each of these cases fall along different parts of the continuum in $\e(R)$ shown in figure \@ref(fig:mu-hyp).

Consider the scenario where the average in resource abundance, $\e(R)$, changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (see figure \@ref(fig:multipanel-constant-var)a). We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b). Since $R$ is spatiotemporally random, an organism that moves in the landscape will not find the same $R$ at different time points or throughout the habitat. However, if $\mu(t)$ changes repetitively and regularly over time (e.g. peaks each spring and is lowest in winter), an organism may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017], but it will not be able to predict *where* $R$ is high if $\e(R)$ does not vary predictably over space.

@@

```{r s2-hyp, fig.cap="Hypothesized space-use requirements of an organism as a function of resource stochasticity ($\\var(R)$). We expect low values of $\\var(R)$ to result in low space-use requirements as organisms are able to depend on relatively predictable resources. As $\\var(R)$ increases, space-use requirements should increase nonlinearly, whether this results in an expansion of the home range (in the case of range-resident organisms) or a switch to dispersal, nomadism, or migration. Note that the relationship between $\\var(R)$ and space-use requirements cannot be linear because it would require space-use requirements to be negative for low values of $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/variance-abundance-hr-hypotheses.png')
```

@@

```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with changing variance (even after accounting for changes in the mean). (a.) Although resources ($R$, brown-green fill) varies over time and space, the variance (dot color) is lowest at the beginning and end of the observational period and highest when $\\e(R)$ peaks. (b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and variance $\\sigma^2(t)$. The points in the central panel indicate the ralizations of $R$ in panel a.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/habitat-examples-changing-variance.png')
```

Similar to the effect of $\e(R)$, we expect $\var(R)$ to have a nonlinear effect on organisms' space-use requirements. When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect, but as resources become increasingly unpredictable, we expect space-use requirements to increase progressively faster (figure \@ref(fig:s2-hyp)) because: (1) as $\var(R)$ increases the chances of finding low $R$ increase superlinearly, (2) stochastic environments tend to be less productive [@chevin_adaptation_2010], and (3) the added movement required to search for food increases organisms' energetic requirements. If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), organisms may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage or food caching). If changes in $\sigma^2(t)$ are sufficiently predictable, organisms may learn to anticipate and prepare for periods of greater stochasticity by pre-preemptively caching food or migrating, or relying on alternative food sources during stochastic events (e.g., after the arrival of competitors).

>>>>

---
title: \large How resource abundance and stochasticity affect organisms' space-use requirements
author:
  - name: Stefano Mezzini
    institute: [braes, biol]
  - name: Simon Wood
    institute: uofe
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: E. Patrícia Medici
    institute: [tapir, iucn, escas]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
    correspondence: "yes"
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - uofe: School of Mathematics, James Clerk Maxwell Building, University of Edinburgh, Edinburgh, Unted Kingdom.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'hr-resource-stoch.bib'
csl: 'the-american-naturalist.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{wrapfig} # for wrapping text around figures
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # @reed_frequency_2003 has unicode character U+0308
    pandoc_args:
# https://github.com/pandoc/lua-filters/blob/master/author-info-blocks/README.md
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
editor_options: 
  chunk_output_type: console
---

@@

<!-- place authors' emails on a new page -->

\noindent \textbf{Authors' emails:}

\noindent Stefano Mezzini: stefano.mezzini@ubc.ca

\noindent Simon Wood: simon.wood@bath.edu

\noindent Chris H. Fleming: chris.h.fleming@gmail.com

\noindent E. Patrícia Medici: medici@ipe.org.br

\noindent Michael J. Noonan: michael.noonan@ubc.ca

@@

knitr::knit_hooks$set(plot = function(x, options) { # set the new plot hook...

@@

  # if wrapfigure options are given, create the new opening string for the
  # wrapfigure environment using custom figure alignment and text width
  if(!is.null(options$wrapfigure)) {

@@

\linenumbers

@@

\noindent The amount of space organisms use is thought to be tightly linked to the availability of resources within their habitats, such that organisms living in productive habitats generally require less space than those in resource-poor habitats. This hypothesis has widespread empirical support, but existing studies have focused primarily on responses to the *mean* amount of resources, while responses to the variance around the mean are still largely unknown. This is not a trivial oversight. Organisms adjust to variable environmental conditions, so failing to consider the effects of resource (un)predictability can result in a limited understanding of organisms' space-use requirements, challenging ecological theory and applied conservation alike. In this study, we first review the literature of studies that estimate organisms' space-use requirements as a function of resource abundance. We then leverage this information to provide a unifying framework and hypotheses for the effect of mean *and* variance in resources on organisms' space use. Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resource abundance interact to shape predictable patterns in space use. Finally, we use real-world tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals. Results from the simulations and empirical examples are presented using a fully transparent approach that allows researchers to apply the framework to their own data and inform area-based conservation efforts.

\vspace*{\fill}

\noindent \textbf{Key words:} energetics, energetic landscape, environmental stochasticity, space use, spatial needs, `ctmm`.

@@

The relationship between organisms' movement and resource abundance has been of interest to biologists for nearly a century. @burt_territoriality_1943 considered the search for food as the primary driver for movement within an organism's home range. Three decades after, @southwood_habitat_1977 suggested that change in resource abundance is a strong determinant of how organisms decide where to live and when to reproduce. Two years later, @harestad_home_1979 proposed that the simplest relationship between resource abundance and an organism's home-range size is

@@

\noindent where $H$ is the organism's **home-range size**, $C$ is the organism's resource consumption (kcal day$^{-1}$), and $R$ is the **resources** the organism can access (kcal day$^{-1}$ unit area$^{-1}$). Harestad and Bunnel's model is simple to conceptualize and allows for testable predictions, but an organism's values of $C$ and $R$ likely depend on numerous factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020]. While many researchers have since demonstrated that organisms adapt their home ranges in response to resources abundance, few studies build upon others' previous work. Instead, results are typically reported as independent, novel findings. Consequently, we currently lack a unifying framework for quantifying the effects of resource abundance on organisms' space-use requirements. In addition, while much work has been done on estimating organisms' responses to average resource abundance, there is little information on how they respond to unpredictable changes in resources.

Here, we refer to a location's average amount of resources as **resource abundance**, while we use the phrase **resource stochasticity** to indicate the variability in resource abundance due to unpredictable causes. We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an organism might use. For instance, a grassland with relatively low but constant forage availability will require drastically different behaviors and adaptations from a desert location with equally scarce forage but rare, sudden, and strong pulses of resources. In the grassland, an animal may require a large but constant home range as it moves between locations in search of food [*sensu* @teitelbaum_beyond_2019], while an animal in the desert may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are depleted. Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in substantially different movement and life history strategies being selected for. Although it is generally expected that resource unpredictability will decrease organisms' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is limited empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].

Here, we illustrate how organisms' space use depends on both the abundance and unpredictability of resources. First, we set the theoretical background necessary for the successive sections by introducing key concepts and notation. Next, we provide a review of the effects of resource *abundance* on organisms' space use while suggesting a unifying hypothesis. Afterwards, we provide a review of the effects of resource *stochasticity* on organisms' space use while suggesting a second unifying hypothesis. Subsequently, we demonstrate the power of these two hypotheses using quantitative, simulated responses to changes in resource abundance and unpredictability. Finally, we demonstrate how this framework can be used in practice to describe the movement ecology of a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado [@medici_movement_2022]. Results from the simulations and empirical example are presented using a fully transparent approach that allows researchers to replicate the work and apply the methods to their own tracking data.

# Resources as a random variable

\noindent In statistics, **random variables** indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$). Known values, such as realizations of random variables (i.e., known observations or instances), are indicated with lower-case letters (e.g., $r$). Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$. Resources are often unpredictable (and difficult to quantify), since they depend on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at the individual [@schmidt_interplay_2020] and species level [@jetz_scaling_2004]. Thus, we can let the random variable $R$ indicate the amount of resources at a given point in space and time. Quantifying resources as a numerical random variable, as opposed to using *ad hoc* qualitative descriptions, provides us with the capacity to leverage techniques from probability theory and statistics.

In this section, we simulate $R$ using a Gamma distribution with time-dependent (but spatially homogeneous) mean $\mu(t)$ and variance $\sigma^2(t)$, which we write as $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$. Although Gamma distributions are more often parameterized using parameters shape and scale ($k > 0$ and $\theta > 0$) or shape and rate ($\alpha > 0$ and $\beta = 1/\theta > 0$), we use $\e(R) = k \theta$ and $\var(R) = k \theta^2$ to facilitate visualizing the simulations. However, note that $\e(R)$ and $\var(R)$ are not independent because the variance depends strongly on the mean (and vice-versa). As the mean approaches zero (from the positive side, which we write as $\e(R) \rightarrow0^+$) the variance also does: $\e(R) \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \var(R) = 0^+$. This assumption also holds biologically, since resources tend to be less variable when they are less abundant. We assume readers are familiar with the concepts of: random variables, probability distributions, expected value of a random variable, and variance of a random variable, but we define and explain each of the four concepts in Appendix 1.

@@

\noindent While organisms' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction. All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001]. Failing to acquire sufficient resources can result in lower fitness, behavioral changes, physiological changes or damage, inability to reproduce, and death. Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism and movement costs [@taylor_energetics_1982]. The hypothesis that space-use requirements decrease with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014]. This is problematic because a linear relationship can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022]. Conceptually, the relationship between space-use requirements and $\e(R)$ must be nonlinear, since there are limits to how much space an organism is able to explore in its lifetime as well as the minimum amount of space it requires to survive [which is necessarily greater than zero -- see fig. \@ref(fig:mu-hyp) as well as: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020; and contrast them to estimates based on linear models, e.g.: @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022].

```{r mu-hyp, fig.cap="Hypothesized space-use requirements of an organism as a function of resource abundance ($\\e(R)$). We expect low values of $\\e(R)$ to result in high space-use requirements as organisms are forced to explore large areas to collect the resources they require to survive, whether they be range-resident or nomadic. As $\\e(R)$ increases, space-use requirements should decrease nonlinearly until they reach the minimum amount of space required by the organism to survive. Note that the relationship between $\\e(R)$ and space-use requirements cannot be linear because it would require space-use requirements to be negative for high values of $\\e(R)$.", out.width = '0.5\\textwidth'}
knitr::include_graphics('../figures/mean-abundance-hr-hypotheses.png')
```

@@

It is also worth noting that high $\e(R)$ does not does not necessarily imply that organisms have access to high amounts of resources, as predator avoidance and competition within an organism's habitat will decrease the proportion of $R$ it is able to access. Thus, organisms in habitats with strong competitive pressure and predation may require larger home ranges [@jetz_scaling_2004; @prox_framework_2020], but larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004; but also see @dickie_resource_2022]. Additionally, persistent territorial defense from competitors may prevent organisms from using space freely and as necessary [wolves, *Canis lupus*: @rich_anthropogenic_2012; feral cats, *Felis catus*: @bengsen_feral_2016; Capuchin monkeys, *Cebus capucinus*: @torrez-herrera_monkeys_2020], so the effect of $\e(R)$ on space use may vary between individuals, species, and locations. Still, these pressures simply alter the intensity with which organisms respond to $\e(R)$ but not the shape of the relationship, since each of these cases fall along different parts of the continuum in $\e(R)$ shown in fig. \@ref(fig:mu-hyp).

```{r multipanel-constant-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with constant variance (after accounting for changes in the mean). (a.) Resources ($R$, brown-green fill) vary over time and space with a constant the variance (dot color). (b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and constant variance $\\var(R) = \\sigma^2$. The points in the central panel indicate the ralizations of $R$ in panel a.", out.width = '0.5\\textwidth', fig.pos='!t'}
knitr::include_graphics('../figures/habitat-examples-constant-variance.png')
```

Consider the scenario where the average in resource abundance, $\e(R)$, changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (see fig. \@ref(fig:multipanel-constant-var)a). We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (fig. \@ref(fig:multipanel-constant-var)b). Since $R$ is spatiotemporally random, an organism that moves in the landscape will not find the same $R$ at different time points or throughout the habitat. However, if $\mu(t)$ changes repetitively and regularly over time (e.g. peaks each spring and is lowest in winter), an organism may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017], but it will not be able to predict *where* $R$ is high if $\e(R)$ does not vary predictably over space.

@@

```{r s2-hyp, fig.cap="Hypothesized space-use requirements of an organism as a function of resource stochasticity ($\\var(R)$). We expect low values of $\\var(R)$ to result in low space-use requirements as organisms are able to depend on relatively predictable resources. As $\\var(R)$ increases, space-use requirements should increase nonlinearly, whether this results in an expansion of the home range (in the case of range-resident organisms) or a switch to dispersal, nomadism, or migration. Note that the relationship between $\\var(R)$ and space-use requirements cannot be linear because it would require space-use requirements to be negative for low values of $\\var(R)$.", out.width = '0.5\\textwidth'}
knitr::include_graphics('../figures/variance-abundance-hr-hypotheses.png')
```

@@

Similar to the effect of $\e(R)$, we expect $\var(R)$ to have a nonlinear effect on organisms' space-use requirements. When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect, but as resources become increasingly unpredictable, we expect space-use requirements to increase progressively faster (fig. \@ref(fig:s2-hyp)) because: (1) as $\var(R)$ increases the chances of finding low $R$ increase superlinearly, (2) stochastic environments tend to be less productive [@chevin_adaptation_2010], and (3) the added movement required to search for food increases organisms' energetic requirements. If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), organisms may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage or food caching). If changes in $\sigma^2(t)$ are sufficiently predictable, organisms may learn to anticipate and prepare for periods of greater stochasticity by pre-preemptively caching food or migrating, or relying on alternative food sources during stochastic events (e.g., after the arrival of competitors).

Like the scenario illustrated in fig. \@ref(fig:multipanel-constant-var), we can imagine an environment where both $\e(R)$ and $\var(R)$ change over time. We can then define $\var(R)$ as a function of time using the same notation as we did for $\e(R)$, i.e., $\var(R | t) = \sigma^2(t)$ (fig. \@ref(fig:multipanel-changing-var)). When both $\e(R)$ and $\var(R)$ change over time, predicting $R$ becomes more complex. Statistically, this is because $\sigma^2(t)$ decreases as we allow $\mu(t)$ to change more over time and explain a greater proportion of $\var(R)$. Biologically, this is important because some organisms may perceive changes in $R$ as a change in $\e(R)$, while others may perceive the change as a stochastic event. An organism's perception of changes in $R$ will depend strongly on the organism's cognitive capacities, including its memory and lifespan [@mueller_social_2013; @abrahms_memory_2019; @foley_severe_2008; @falcon-cortes_hierarchical_2021; @fagan_spatial_2013]. The ability to predict trends in $\mu(t)$ and $\sigma^2(t)$ requires organisms have high cognitive capacity, since distinguishing between changes in $\e(R)$ and $\var (R)$ is not easy [@steixner-kumar_strategies_2020], especially if $\e(R)$ and $\var(R)$ are not independent, as in the case of $R \sim \Gamma(\mu(t), \sigma^2(t))$. Regardless of whether an organism is capable of predicting changes in $\sigma^2(t)$, the effect of $\var(R)$ is arguably as important as the effect of $\e(R)$ in determining how said organism will respond to changes in $R$ and consequently adapt its space use.

```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with changing variance (even after accounting for changes in the mean). (a.) Although resources ($R$, brown-green fill) varies over time and space, the variance (dot color) is lowest at the beginning and end of the observational period and highest when $\\e(R)$ peaks. (b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and variance $\\sigma^2(t)$. The points in the central panel indicate the ralizations of $R$ in panel a.", out.width = '0.5\\textwidth', fig.pos='!t'}
knitr::include_graphics('../figures/habitat-examples-changing-variance.png')
```
